A highly-available move operation for replicated trees

### 2019-06-18 paper/move-op.tex
paper: better reference for Coda
paper: tidy up discussion of hardlinks

For many systems and applications, the data model that describes the system state is a tree.
For example:
\begin{itemize}
    \item Most \textbf{filesystems}, e.g. on Windows and Unix-like OSes, are organised as a tree: folders or directories are branch nodes, and files are leaf nodes.\footnote{In filesystems that support hardlinks, the structure is---strictly speaking---a restricted DAG in which a file inode can be referenced from multiple places in the directory tree, but directories cannot have more than one parent directory.
        However, for the purposes of this paper, we will treat a filesystem as a tree; given such a tree, it is easy to add hardlinks and symlinks as features on top.}
        In Unix-style filesystems, the nodes of the tree are known as \emph{inodes}.
    \item \textbf{Rich text editors} maintain a tree of textual structures such as paragraphs, bulleted and numbered lists, figures, sections, and so on.
        These can be nested inside each other: for example, a paragraph may appear inside a bullet point of a bulleted list, which may in turn appear inside a numbered point of an enumeration.
        Rich text documents are typically represented in memory using the Document Object Model (DOM), and written to file in HTML, XML, or JSON format.
    \item \textbf{Vector graphics} and \textbf{presentation software} represents images using graphical objects such as text boxes, rectangles, ellipses, lines, and so on.
        These objects are contained within nodes representing pages of a document, or slides of a presentation.
        Multiple objects may be combined into a \emph{group} so that they can be manipulated as a unit; this corresponds to making these objects children of a common parent node.
        Multiple objects and groups may in turn be combined into higher-level groups, forming a tree.
    \item Personal \textbf{note-taking} and \textbf{task-management tools} such as Org-mode for Emacs~\cite{OrgMode} or OmniOutliner~\cite{OmniOutliner} explicitly present the user with a tree structure that they can inspect and manipulate.
\end{itemize}

@@

The algorithm as specified in Figure~\ref{fig:code} retains operations in the log indefinitely, so the memory/storage use grows without bound.
However, in practice it is easy to truncate the log, because \isa{apply\_op} only examines the prefix of operations whose timestamp is greater than that of the operation being applied.
Thus, once it is known that all future operations will have a timestamp greater than $t$, then all operations with timestamp $t$ or less can be discarded from the log.
In this case, we say that $t$ is \emph{causally stable}~\cite{Baquero:2014ed}.

@@

@@

\noindent Let \isa{t1 = move\_time oper1} and \isa{t2 = move\_time oper2} be the timestamps of the two operations \isa{oper1} and \isa{oper2}.
Assume \isa{t1 < t2} without loss of generality.
We can now prove lemma \isa{apply\_op\_commute2} by induction over \isa{log}.

@@

TODO: Coda~\cite{Kumar:1995wf}, Ficus~\cite{Reiher:1994wh}, Rumor~\cite{Guy:1999gy}, Roam~\cite{Ratner:1999fh}

@@

>>>>

For many systems and applications, the data model that describes the system state is a tree.
For example:
\begin{itemize}
    \item The \textbf{filesystem} on most operating systems is organised as a tree: folders or directories are branch nodes, and files are leaf nodes.
        In Unix-style filesystems, the nodes of the tree are known as \emph{inodes}.
        Strictly speaking, filesystems that support hardlinks form a DAG, but we will focus on trees for now and show in Section~\ref{sec:hardlinks} how to add support for links.
    \item \textbf{Rich text editors} maintain a tree of textual structures such as paragraphs, bulleted and numbered lists, figures, sections, and so on.
        These can be nested inside each other: for example, a paragraph may appear inside a bullet point of a bulleted list, which may in turn appear inside a numbered point of an enumeration.
        Rich text documents are typically represented in memory using the Document Object Model (DOM), and written to file in HTML, XML, or JSON format.
    \item \textbf{Vector graphics} and \textbf{presentation software} represents images using graphical objects such as text boxes, rectangles, ellipses, lines, and so on.
        These objects are contained within nodes representing pages of a document, or slides of a presentation.
        Multiple objects may be combined into a \emph{group} so that they can be manipulated as a unit; this corresponds to making these objects children of a common parent node.
        Multiple objects and groups may in turn be combined into higher-level groups, forming a tree.
    \item Personal \textbf{note-taking} and \textbf{task-management tools} such as Org-mode for Emacs~\cite{OrgMode} or OmniOutliner~\cite{OmniOutliner} explicitly present the user with a tree structure that they can inspect and manipulate.
\end{itemize}

@@

\subsubsection{Hardlinks and symlinks}\label{sec:hardlinks}

Unix filesystems support hardlinks, which allow the same file inode to be referenced from multiple locations in a tree.
Our tree data structure can easily be extended to support this: rather than placing file data directly in the leaf nodes of the tree, the leaf node must reference the file inode.
Thus, references to the same file inode can appear in multiple leaf nodes of the tree.
Symlinks are also easy to support, since they are just leaf nodes containing a path (not a reference to an inode).

\subsubsection{Log truncation}\label{sec:log-truncation}

The algorithm as specified in Figure~\ref{fig:code} retains operations in the log indefinitely, so the memory use grows without bound.
However, in practice it is easy to truncate the log, because \isa{apply\_op} only examines the prefix of operations whose timestamp is greater than that of the operation being applied.
Thus, once it is known that all future operations will have a timestamp greater than $t$, then all operations with timestamp $t$ or less can be discarded from the log.
In this case, we say that $t$ is \emph{causally stable}~\cite{Baquero:2014ed}.

@@

\subsubsection{Ordering of sibling nodes}\label{sec:ordering}

@@

\noindent Here the \isa{\isacharat} operator concatenates two lists.
Let \isa{t1 = move\_time oper1} and \isa{t2 = move\_time oper2} be the timestamps of the two operations \isa{oper1} and \isa{oper2}.
Assume \isa{t1 < t2} without loss of generality.
We can now prove lemma \isa{apply\_op\_commute2} by induction over \isa{log}.

@@

TODO: Coda~\cite{Kumar:1993ey}, Ficus~\cite{Reiher:1994wh}, Rumor~\cite{Guy:1999gy}, Roam~\cite{Ratner:1999fh}

@@

% Disconnected operation
